\vspace{-.6in}Annotation-Based Empirical Performance Tuning\\ Using Orio

### 2009-02-12 doc/papers/ipdps09/conclusions.tex
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@427 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

\vspace{-.1in} 
\paragraph{Acknowledgments.} We would like to thank
Uday Bondhugula of Ohio State University for many productive discussions and
his valuable help with PLuTo.

>>>>

\vspace{-.1in} 
\paragraph{Acknowledgments.} We would like to thank
Uday Bondhugula of Ohio State University for many productive discussions and
his valuable help with Pluto.

### 2009-02-12 doc/papers/ipdps09/integration.tex
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@427 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

\section{PLuTo-Orio Integration}
\label{sec:integration}

A number of source-to-source transformation tools for performance
optimization exist. Using these tools to achieve (near) optimal performance
on different architectures, however, is still a nontrivial task that requires
significant architectural and compiler expertise. By combining code
transformation tools with an empirical tuning system, such as Orio, we can
reduce the amount of manual effort and automatically determine the
transformations that result in the best performance. We show in this section
how Orio has been extended with an external transformation program, called
PLuTo~\cite{uday08pldi}. This integration also demonstrates the easy
extensibility of Orio and the ability to leverage other source transformation
approaches.

PLuTo is a source-to-source automatic transformation tool aimed at optimizing
a sequence of nested loops for data locality and coarse-grained parallelism
simultaneously. PLuTo employs a polyhedral model of arbitrary loop nests,
where the dynamic instance (iteration) of each statement is viewed as an
integer point in a well-defined space called the statement's polyhedron. This
statement representation and a precise characterization of data dependences
enable PLuTo to construct mathematically correct complex loop
transformations. PLuTo's polyhedral-based transformations result in improved
cache locality and loops parallelized for multicore architecture.

Figure~\ref{fig:pluto-orio} outlines the overall structure of the PLuTo-Orio
integrated system, which is implemented as a new optimization module in
Orio.
%% BN: the detailed description below is good, but commented out for now for space.
%Initially, the transformation process takes two kinds of input:
%the loop code to be optimized and the optimization parameters code. The loop
%code is embedded in the annotation body block, while the performance
%parameters code is written in the module body block. A simple parser
%implemented inside the new module parses the performance parameters code to
%extract their values. 
In addition to the loop code to be optimized, the module parses performance
parameters, which include tile sizes, unroll factors, loop order
(permutation), as well as several boolean values for triggering OpenMP
parallelization, scalar replacement, and auto-vectorization. Using these
parameter values, PLuTo then performs polyhedral transformations (e.g.,
two-level tiling and OpenMP parallelization) on the input loop code. The
resulting generated code is subsequently passed to the widely-available
profiling tool gprof~\cite{gprof} for hotspot detection. 
%The statistical
%information produced by gprof provides accurate locations (line numbers) of
%the loop nests where the PLuTo-generated code spends most of its execution
%time. 
The identified hotspot loop nests are then decorated with Orio's performance
annotations for complementary syntactic transformations (e.g., loop
permutation, loop unroll/jamming, scalar replacement, and explicit
auto-vectorization). Finally, Orio optimizes the annotated hotspots as 
described in Section~\ref{sec:implm}.

\begin{figure}[tb]
\begin{center}
\includegraphics[width=0.6\textwidth]{figures/pluto-orio.eps}    
\end{center}   
\vspace{-.2in}
\caption{Integration of PLuTo into Orio's code transformation module.}
\label{fig:pluto-orio}
\end{figure}

PLuTo also performs syntactic loop unroll/jamming as a post-processing
pass. The target loops, however, are limited only to innermost loops with a
maximum depth of two (i.e., 1-D unrolling and 2-D unroll/jamming). So we
choose to use the loop unroll/jamming transformation already available in
Orio, which is more flexible because it can be applied to loop nests of
depths larger than two.

At present Orio does not employ any data dependency analysis when performing
syntactic transformations. Therefore, there is no guarantee that the code
generated after syntactic transformations is correct. Because of this, the
tuning process using the integrated PLuTo-Orio system is currently
semi-automatic---user involvement is required to decide whether it is safe
to apply a syntactic transformation on an identified hotspot.
%The user only needs
%to check the transformation legalities once, since the hotspots
%consistently have similar basic loop structure for different
%PLuTo-generated code versions. Once all illegal syntactic
%transformations are known by the user, they must be switched off so
%that the tuning process can proceed fully automatic with no more human
%interventions.

>>>>

\section{Pluto-Orio Integration}
\label{sec:integration}

A number of source-to-source transformation tools for performance
optimization exist. Using these tools to achieve (near) optimal performance
on different architectures, however, is still a nontrivial task that requires
significant architectural and compiler expertise. By combining code
transformation tools with an empirical tuning system, such as Orio, we can
reduce the amount of manual effort and automatically determine the
transformations that result in the best performance. We show in this section
how Orio has been extended with an external transformation program, called
Pluto~\cite{uday08pldi}. This integration also demonstrates the easy
extensibility of Orio and the ability to leverage other source transformation
approaches.

Pluto is a source-to-source automatic transformation tool aimed at optimizing
a sequence of nested loops for data locality and coarse-grained parallelism
simultaneously. Pluto employs a polyhedral model of arbitrary loop nests,
where the dynamic instance (iteration) of each statement is viewed as an
integer point in a well-defined space called the statement's polyhedron. This
statement representation and a precise characterization of data dependences
enable Pluto to construct mathematically correct complex loop
transformations. Pluto's polyhedral-based transformations result in improved
cache locality and loops parallelized for multicore architecture.

Figure~\ref{fig:pluto-orio} outlines the overall structure of the Pluto-Orio
integrated system, which is implemented as a new optimization module in
Orio.
%% BN: the detailed description below is good, but commented out for now for space.
%Initially, the transformation process takes two kinds of input:
%the loop code to be optimized and the optimization parameters code. The loop
%code is embedded in the annotation body block, while the performance
%parameters code is written in the module body block. A simple parser
%implemented inside the new module parses the performance parameters code to
%extract their values. 
In addition to the loop code to be optimized, the module parses performance
parameters, which include tile sizes, unroll factors, loop order
(permutation), as well as several boolean values for triggering OpenMP
parallelization, scalar replacement, and auto-vectorization. Using these
parameter values, Pluto then performs polyhedral transformations (e.g.,
two-level tiling and OpenMP parallelization) on the input loop code. The
resulting generated code is subsequently passed to the widely-available
profiling tool gprof~\cite{gprof} for hotspot detection. 
%The statistical
%information produced by gprof provides accurate locations (line numbers) of
%the loop nests where the Pluto-generated code spends most of its execution
%time. 
The identified hotspot loop nests are then decorated with Orio's performance
annotations for complementary syntactic transformations (e.g., loop
permutation, loop unroll/jamming, scalar replacement, and explicit
auto-vectorization). Finally, Orio optimizes the annotated hotspots as 
described in Section~\ref{sec:implm}.

\begin{figure}[tb]
\begin{center}
\includegraphics[width=0.6\textwidth]{figures/pluto-orio.eps}    
\end{center}   
\vspace{-.2in}
\caption{Integration of Pluto into Orio's code transformation module.}
\label{fig:pluto-orio}
\end{figure}

Pluto also performs syntactic loop unroll/jamming as a post-processing
pass. The target loops, however, are limited only to innermost loops with a
maximum depth of two (i.e., 1-D unrolling and 2-D unroll/jamming). So we
choose to use the loop unroll/jamming transformation already available in
Orio, which is more flexible because it can be applied to loop nests of
depths larger than two.

At present Orio does not employ any data dependency analysis when performing
syntactic transformations. Therefore, there is no guarantee that the code
generated after syntactic transformations is correct. Because of this, the
tuning process using the integrated Pluto-Orio system is currently
semi-automatic---user involvement is required to decide whether it is safe
to apply a syntactic transformation on an identified hotspot.
%The user only needs
%to check the transformation legalities once, since the hotspots
%consistently have similar basic loop structure for different
%Pluto-generated code versions. Once all illegal syntactic
%transformations are known by the user, they must be switched off so
%that the tuning process can proceed fully automatic with no more human
%interventions.

### 2009-02-12 doc/papers/ipdps09/results.tex
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@364 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@427 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

\subsection{Evaluation of the PLuTo-Orio Integrated System}

This section discusses the performance evaluation of the integrated
PLuTo-Orio system (discussed in Section~\ref{sec:integration}) on the
multicore Intel Xeon by using a number of application kernels that are
nontrivial to optimize and parallelize. We compare the performance of the
code tuned by Orio with the base code and the PLuTo-generated codes. The
PLuTo code was obtained by running the base code with
PLuTo-0.3.0~\cite{pluto030} using \texttt{--tile} to employ loop tiling for
L1 cache, \texttt{--l2tile} to employ loop tiling for L2 cache, and
\texttt{--unroll} to employ loop unrolling, and for parallel code generation,
an additional \texttt{--parallel} option. We used PLuTo's default tile
sizes (L1: 32x32 or 32x32x32; L2: 256x256 or 256x256x256) and default
unroll factors (64 for 1-D unrolling and 8x8 for 2-D
unroll/jamming). All codes were compiled with the Intel C Compiler
using the \texttt{-O3} optimization flag to enable auto-vectorization
and other advanced optimizations, and
\texttt{-parallel} (or \texttt{-openmp} in the case of manual OpenMP parallelization) 
to enable code parallelization.

In the following sections, we refer to the icc-optimized base code as
``ICC,'' the PLuTo-generated code with L1 tiling and
unroll/jamming as ``PLuTo (L1 tiling),'' and the PLuTo-generated code with L1
and L2 tilings and unroll/jamming as ``PLuTo (L1+L2 tiling).'' The code
tuned by Orio is referred to as ``PLuTo+Orio.''

@@

The performance of the sequential 2-D FDTD code for $tmax=500$ and $nx=ny$ is
shown in Figure~\ref{fig:fdtd-2d-cookie-seq}. The base code optimized by icc
alone performs better than PLuTo for small problem sizes since all input
arrays fit in the L2 cache (insufficient computation to offset PLuTo's tiling
overhead). As the array sizes increase, the lack of data reuse impairs the
base code's performance, whereas the PLuTo performance remains about the
same.
%We used Orio to generate two code variants: one for small problem sizes
%and another for large problem sizes. 
When the input arrays are small, Orio discovers that applying
PLuTo's polyhedral transformations is not beneficial, and therefore,
it employs only its syntactic transformations on the original FDTD
code. For large problem sizes, Orio exploits some of the PLuTo's code
transformations and enhances it further with its syntactic
optimizations, resulting in performance consistently and significantly
higher than both the base and PLuTo codes (up to 86\% over PLuTo).

%By tuning the sequential code with Orio,
%the attainable performance enhancement over PLuTo is up to 86\%.

Figure~\ref{fig:fdtd-2d-cookie-par} shows the multicore performance obtained
for $tmax=500$ and $nx=ny=2000$. The results indicate that whereas icc is
unable to auto-parallelize the code, PLuTo detects the existence of pipelined
parallelism and then successfully parallelizes the code. Optimizing and
tuning by Orio additionally improves the PLuTo performance by up to
78\%. Moreover, we observe that because of memory contention between the two
quad-core Intel processors, the speedup of both the PLuTo and Orio codes
slightly decreases when the number of cores used is greater than four.

@@

%%BN: commented out since this needs more explanation (no room in this version)
%It is to be noted that collecting performance results of the PLuTo 
%code with L1 and L2 tilings was not possible because of the large code 
%size of the generated loop nest, which led to an impractically huge 
%compile time.

%% BN: comment for space
\comment{
\subsubsection{3-D Gauss-Seidel Successive Over-Relaxation Method}  
Figure~\ref{fig:seidel-code} shows the 3-D Gauss-Seidel computation, 
which is sometimes is referred to as
\emph{successive displacement method}, indicating the dependence of the 
iterations on the ordering. If the ordering is
%
%\begin{figure}[ht]
\begin{wrapfigure}{r}{3in}
\vspace{-0.3in}
\begin{center}
\begin{minipage}{3in} 
\scriptsize
\begin{verbatim} 
for (t=0; t<=T-1; t++) 
  for (i=1; i<=N-2; i++) 
    for (j=1; j<=N-2; j++) 
      A[i][j] = (A[i-1][j-1] + A[i-1][j] 
          + A[i-1][j+1] + A[i][j-1] + A[i][j] 
          + A[i][j+1] + A[i+1][j-1] + A[i+1][j] 
          + A[i+1][j+1]) / 9.0; 
\end{verbatim} 
\end{minipage} 
\end{center}
\vspace{-0.2in}
\caption{3-D Gauss-Seidel code} 
\label{fig:seidel-code} 
\vspace{-.1in}
\end{wrapfigure}
%
changed, the components of the new iterations will change as well.

Figure~\ref{fig:seidel-cookie-seq} contains the sequential performance
results for $T=500$, which show that applying PLuTo's polyhedral tiling on
the original code always delivers performance boosts, which range from 126\%
to 142\%. When using PLuTo, performing two-level tiling (for both L1 and L2
caches) yields 13\% lower performance than performing only L1 tiling. 
This is also reflected in the best sequential code found by Orio, where
one-level tiling (for L1 cache only) is always performed. The sequential
speedup obtained from tuning the PLuTo code with Orio is significant, ranging
from 51\% to 133\%.

@@

The parallel performance results for $T=500$ and $N=4000$ are shown in
Figure~\ref{fig:seidel-cookie-par}. Again we observe that icc failed to
parallelize the code, whereas PLuTo was able to parallelize it and Orio
improved the performance of the one-level tiled PLuTo code further by up to
128\% (and up to 548\% for the two-level tiled PLuTo code). We observed that
tiling over both L1 and L2 can result in worse performance than L1-only
tiling because the number of tiles can be smaller than the number of cores.
%Similar earlier trend also exists
%in these performance results: PLuTo parallelizes the code through its
%polyhedral transformations (i.e., skewing and tiling), while the icc
%compiler alone does not parallelize the loop. Empirical tuning via
%Orio creates up to 122\% speedups over the one-level tiled PLuTo code,
%and up to 548\% speedups over the two-level tiled PLuTo code. Another
%finding is that tiling for both L1 and L2 caches contributes to a poor
%scalability because the number of tiles can be less than the number of
%processors, resulting in underutilization of the available resources.
} % end comment

@@

The sequential performance results are shown in
Figure~\ref{fig:lu-cookie-seq}. Similarly to the 2-D FDTD results, the icc-optimized code is more
efficient than the PLuTo-generated codes when the input arrays are small and
fit in the L2 cache. Because of better data locality, however, the
performance of the PLuTo-tiled codes is better than that of the base code as
the input arrays get larger. 
%Again, there are two different codes generated
%by Orio: one for small problem sizes and one for large problem sizes. And
Consequently, Orio employs PLuTo's polyhedral transformations only for large
input sizes prior to applying its own syntactic transformations. 
%Unlike 3-D
%Gauss-Seidel, this time Orio chooses to apply both L1 and L2 tilings when
%generating the best tuned code. 
Compared to the two PLuTo codes, the
Orio-tuned code yields performance improvements ranging between 26\% and 277\%.

The parallel performance results shown in Figure~\ref{fig:lu-cookie-par} were
obtained for $N=4000$. We observe that icc is not able to parallelize the
code, whereas PLuTo achieves higher performance by exploiting multicore
parallelism. Furthermore, both PLuTo-generated codes have comparable
performance, with slightly better scalability exhibited by the single-level tiled
code. Orio further improved the performance of the PLuTo-generated versions
by a factor of 51\% to 120\%.

>>>>

\subsection{Evaluation of the Pluto-Orio Integrated System}

This section discusses the performance evaluation of the integrated
Pluto-Orio system (discussed in Section~\ref{sec:integration}) on the
multicore Intel Xeon by using a number of application kernels that are
nontrivial to optimize and parallelize. We compare the performance of the
code tuned by Orio with the base code and the Pluto-generated codes. The
Pluto code was obtained by running the base code with
Pluto-0.3.0~\cite{pluto030} using \texttt{--tile} to employ loop tiling for
L1 cache, \texttt{--l2tile} to employ loop tiling for L2 cache, and
\texttt{--unroll} to employ loop unrolling, and for parallel code generation,
an additional \texttt{--parallel} option. We used Pluto's default tile
sizes (L1: 32x32 or 32x32x32; L2: 256x256 or 256x256x256) and default
unroll factors (64 for 1-D unrolling and 8x8 for 2-D
unroll/jamming). All codes were compiled with the Intel C Compiler
using the \texttt{-O3} optimization flag to enable auto-vectorization
and other advanced optimizations, and
\texttt{-parallel} (or \texttt{-openmp} in the case of manual OpenMP parallelization) 
to enable code parallelization.

In the following sections, we refer to the icc-optimized base code as
``ICC,'' the Pluto-generated code with L1 tiling and
unroll/jamming as ``Pluto (L1 tiling),'' and the Pluto-generated code with L1
and L2 tilings and unroll/jamming as ``Pluto (L1+L2 tiling).'' The code
tuned by Orio is referred to as ``Pluto+Orio.''

@@

The performance of the sequential 2-D FDTD code for $tmax=500$ and $nx=ny$ is
shown in Figure~\ref{fig:fdtd-2d-cookie-seq}. The base code optimized by icc
alone performs better than Pluto for small problem sizes since all input
arrays fit in the L2 cache (insufficient computation to offset Pluto's tiling
overhead). As the array sizes increase, the lack of data reuse impairs the
base code's performance, whereas the Pluto performance remains about the
same.
%We used Orio to generate two code variants: one for small problem sizes
%and another for large problem sizes. 
When the input arrays are small, Orio discovers that applying
Pluto's polyhedral transformations is not beneficial, and therefore,
it employs only its syntactic transformations on the original FDTD
code. For large problem sizes, Orio exploits some of the Pluto's code
transformations and enhances it further with its syntactic
optimizations, resulting in performance consistently and significantly
higher than both the base and Pluto codes (up to 86\% over Pluto).

%By tuning the sequential code with Orio,
%the attainable performance enhancement over Pluto is up to 86\%.

Figure~\ref{fig:fdtd-2d-cookie-par} shows the multicore performance obtained
for $tmax=500$ and $nx=ny=2000$. The results indicate that whereas icc is
unable to auto-parallelize the code, Pluto detects the existence of pipelined
parallelism and then successfully parallelizes the code. Optimizing and
tuning by Orio additionally improves the Pluto performance by up to
78\%. Moreover, we observe that because of memory contention between the two
quad-core Intel processors, the speedup of both the Pluto and Orio codes
slightly decreases when the number of cores used is greater than four.

@@

%%BN: commented out since this needs more explanation (no room in this version)
%It is to be noted that collecting performance results of the Pluto 
%code with L1 and L2 tilings was not possible because of the large code 
%size of the generated loop nest, which led to an impractically huge 
%compile time.

\subsubsection{3-D Gauss-Seidel Successive Over-Relaxation Method}  
Figure~\ref{fig:seidel-code} shows the 3-D Gauss-Seidel computation, 
which is sometimes is referred to as
\emph{successive displacement method}, indicating the dependence of the 
iterations on the ordering. If the ordering is
%
%\begin{figure}[ht]
\begin{wrapfigure}{r}{3in}
\vspace{-0.3in}
\begin{center}
\begin{minipage}{3in} 
\scriptsize
\begin{verbatim} 
for (t=0; t<=T-1; t++) 
  for (i=1; i<=N-2; i++) 
    for (j=1; j<=N-2; j++) 
      A[i][j] = (A[i-1][j-1] + A[i-1][j] 
          + A[i-1][j+1] + A[i][j-1] + A[i][j] 
          + A[i][j+1] + A[i+1][j-1] + A[i+1][j] 
          + A[i+1][j+1]) / 9.0; 
\end{verbatim} 
\end{minipage} 
\end{center}
\vspace{-0.2in}
\caption{3-D Gauss-Seidel code} 
\label{fig:seidel-code} 
\vspace{-.1in}
\end{wrapfigure}
%
changed, the components of the new iterations will change as well.

Figure~\ref{fig:seidel-cookie-seq} contains the sequential performance
results for $T=500$, which show that applying Pluto's polyhedral tiling on
the original code always delivers performance boosts, which range from 126\%
to 142\%. When using Pluto, performing two-level tiling (for both L1 and L2
caches) yields 13\% lower performance than performing only L1 tiling. 
This is also reflected in the best sequential code found by Orio, where
one-level tiling (for L1 cache only) is always performed. The sequential
speedup obtained from tuning the Pluto code with Orio is significant, ranging
from 51\% to 133\%.

@@

The parallel performance results for $T=500$ and $N=4000$ are shown in
Figure~\ref{fig:seidel-cookie-par}. Again we observe that icc failed to
parallelize the code, whereas Pluto was able to parallelize it and Orio
improved the performance of the one-level tiled Pluto code further by up to
128\% (and up to 548\% for the two-level tiled Pluto code). We observed that
tiling over both L1 and L2 can result in worse performance than L1-only
tiling because the number of tiles can be smaller than the number of cores.
%Similar earlier trend also exists
%in these performance results: Pluto parallelizes the code through its
%polyhedral transformations (i.e., skewing and tiling), while the icc
%compiler alone does not parallelize the loop. Empirical tuning via
%Orio creates up to 122\% speedups over the one-level tiled Pluto code,
%and up to 548\% speedups over the two-level tiled Pluto code. Another
%finding is that tiling for both L1 and L2 caches contributes to a poor
%scalability because the number of tiles can be less than the number of
%processors, resulting in underutilization of the available resources.

@@

The sequential performance results are shown in
Figure~\ref{fig:lu-cookie-seq}. Similarly to the 2-D FDTD results, the icc-optimized code is more
efficient than the Pluto-generated codes when the input arrays are small and
fit in the L2 cache. Because of better data locality, however, the
performance of the Pluto-tiled codes is better than that of the base code as
the input arrays get larger. 
%Again, there are two different codes generated
%by Orio: one for small problem sizes and one for large problem sizes. And
Consequently, Orio employs Pluto's polyhedral transformations only for large
input sizes prior to applying its own syntactic transformations. 
%Unlike 3-D
%Gauss-Seidel, this time Orio chooses to apply both L1 and L2 tilings when
%generating the best tuned code. 
Compared to the two Pluto codes, the
Orio-tuned code yields performance improvements ranging between 26\% and 277\%.

The parallel performance results shown in Figure~\ref{fig:lu-cookie-par} were
obtained for $N=4000$. We observe that icc is not able to parallelize the
code, whereas Pluto achieves higher performance by exploiting multicore
parallelism. Furthermore, both Pluto-generated codes have comparable
performance, with slightly better scalability exhibited by the single-level tiled
code. Orio further improved the performance of the Pluto-generated versions
by a factor of 51\% to 120\%.
